---
title: "Two weeks as a MEDS student"
description: |
  A few functions I learned and key takeaways from my first two weeks in MEDS
author:
  - name: Felicia Cruz
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
preview: images/bren_outside.jpeg
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(palmerpenguins)
penguins <- penguins
```

Some functions I have found particularly useful are `group_by()` and `summarize()`. Working together, these two functions will recognize groups that are specified in the arguments and then compute chosen summary statistics represented in a new data frame. 

Here is an example of how these functions can be used to find the mean flipper length for each penguin species in the `palmerpenguins` dataset. 

```{r}
flipper_length_by_species <- penguins %>%
  group_by(species) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE))

flipper_length_by_species
```
Just by using these two functions, from the new data frame we can quickly see that Gentoos have the longest flipper length on average. 

Beyond just learning useful functions, I have also deepened my understanding of tidy data. While I used to think that "tidy data" simply referred to datasets that were visually clean and organized, I have come to learn that tidy data refers to data that follows a certain structure. The three criteria that must be met for data to be considered "tidy" are: 

- Each variable must be a column
- Each observation must be a row
- Each cell can contain only one value

When data is in tidy form, this allows for easier visualizations and analyses. 

Lastly, a key takeaway I have learned in my first two weeks as a data science student is that collaboration is critical. Whether collaboration means testing functions together or simply spotting an unmatched parentheses, I have really enjoyed the teammwork aspect that comes along with working in this field.   

