---
title: "Using an API to Programmatically Access and Download Data"
categories:
  - packages
  - MEDS
description: |
  Using the `dataRetrieval` package to analyze USGS datasets
author:
  - name: Felicia Cruz
    url: {}
date: 10-20-2021
output:
  distill::distill_article:
    self_contained: false
preview: images/dataretrieval_hex.png.crdownload
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Maybe look at the storms that caused the Montecito mudslides in 2018?? 

During my EDS 213 course titled "Metadata Standards, Data Modeling and Data Semantics" I learned how to use an API to access and download data in R. Using the `dataRetrieval` package to access USGS data, I will be looking at the Ventura River and Santa Paula Creek gauges on October 4th.

One of my favorite parts of Santa Barbara is the beautiful weather.

On the night of October 4th as I was driving home from an evening workout on campus, 

```{r, include = FALSE}
library(dataRetrieval)
library(tidyverse)
library(metajam)
library(here)
```

## Reading in the data 
```{r}




```

## Analysis 
